### 2023-05-04 tex/annexes_autres.tex
Fix spelling

Il est souvent utile voire nécessaire d'accompagner certaines quantités par des informations supplémentaires: pour des données satellites par exemple, la présence de nuages sur le pixel, la qualité de la mesure, ou encore le capteur qui a réalisé la mesure (pour un produit composite).
Ces informations sont différentes pour chaque point de mesure, et doivent donc être stockées comme des variables (\as{ie} des tableaux multi-dimensionnels).
Ce type d'information (que nous appelerons drapeaux) peut souvent être représenté de manière très simple: la présence de nuage peut être indiqué par un booléen (un seul bit), la qualité de la mesure est typiquement un petit entier (2 ou 3 bits), etc.

Cependant, du fait de l'architecture des processeurs, les outils informatiques ne permettent généralement pas de manipuler des variables aussi petites.
Chaque variable drapeau ne peut être traitée et stockée que sur un octet, au minimum.
On gâche alors de l'espace de stockage puisque seuls quelques bits sont utilisés sur les huits.

@@

Par ailleurs, on ne récupère que des signications individuelles. Il pourrait être intéressant de pouvoir extraire un ensemble de drapeaux mutuellement exclusifs.
Par exemple, on voudrait pouvoir récupérer toutes les valeurs de qualité, ou d'algorithme utilisé (plutôt que seulement les pixels correspondant à \emph{un} algorithme, ou \emph{une} valeur de qualité).

Enfin, la documentation de cette nouvelle fonctionalité n'a pas encore été rédigée.

@@

Dans cette section, j'estime le coût carbone engendré par mon travail de thèse.
Commencons par les voyages professionels, typiquement un grand poste émetteur personnel dans les carrières scientifiques.

@@

Je comptabilise également les coûts communs liés au fonctionnement du laboratoire (consommation énérgétique, entretien bâtiment, mails,\dots) qui ont été estimés par le groupe Climaction-LOCEAN pour l'année 2018\footnote{\url{https://climactions.ipsl.fr/groupes-de-travail/locean-climactions/}}.
Individuellement, cela représente environ \qty{1}{\tcarbone} annuelle.

>>>>

Il est souvent utile voire nécessaire d'accompagner certaines quantités par des informations supplémentaires: pour des données satellites par exemple, la présence de nuages sur le pixel, la qualité de la mesure, ou encore le capteur qui a réalisé la mesure (pour un produit composite).
Ces informations sont différentes pour chaque point de mesure, et doivent donc être stockées comme des variables (\as{ie} des tableaux multi-dimensionnels).
Ce type d'information (que nous appellerons drapeaux) peut souvent être représenté de manière très simple: la présence de nuage peut être indiqué par un booléen (un seul bit), la qualité de la mesure est typiquement un petit entier (2 ou 3 bits), etc.

Cependant, du fait de l'architecture des processeurs, les outils informatiques ne permettent généralement pas de manipuler des variables aussi petites.
Chaque variable drapeau ne peut être traitée et stockée que sur un octet, au minimum.
On gâche alors de l'espace de stockage puisque seuls quelques bits sont utilisés sur les huit.

@@

Par ailleurs, on ne récupère que des significations individuelles. Il pourrait être intéressant de pouvoir extraire un ensemble de drapeaux mutuellement exclusifs.
Par exemple, on voudrait pouvoir récupérer toutes les valeurs de qualité, ou d'algorithme utilisé (plutôt que seulement les pixels correspondant à \emph{un} algorithme, ou \emph{une} valeur de qualité).

Enfin, la documentation de cette nouvelle fonctionnalité n'a pas encore été rédigée.

@@

Dans cette section, j'estime le coût carbone engendré par mon travail de thèse.
Commençons par les voyages professionnels, typiquement un grand poste émetteur personnel dans les carrières scientifiques.

@@

Je comptabilise également les coûts communs liés au fonctionnement du laboratoire (consommation énergétique, entretien bâtiment, mails,\dots) qui ont été estimés par le groupe Climaction-LOCEAN pour l'année 2018\footnote{\url{https://climactions.ipsl.fr/groupes-de-travail/locean-climactions/}}.
Individuellement, cela représente environ \qty{1}{\tcarbone} annuelle.

### 2023-05-04 tex/annexes_sms.tex
Fix spelling

\lit{args} is a dictionnary containing arguments name and their values.

\subsection{Default arguments}
The list of arguments names passed above (region, days, fixes) are from a list of default arguments.
Thoses are defined globally in \linkfile{lib/conf.ini}{lib/conf.ini}.
This file must be created; it can be copied from \linkfile{lib/conf_example.ini}{lib/conf_example.ini}.
Each default argument should have a name, type, and value. See the example conf file for more details.
The default values can be overidden when creating a dataset: see \titleref{sec:org-datasets}.

@@

\subsubsection{fixes}
Fixes are supplied as \lit{-fix <matcher> <string>}. They can then be found as a dictionnary with the matchers name as keys. %chktex 44
See \titleref{sec:org-fixes} below.

@@

Now, we can do this automatically from the command line by supplying the appropriate argument: \shell{python script.py -fix m 05} (from the command line we can only supply a string, so it must match the date format). This will end up in the args dictionnary, where it can be passed to \lit{get_finder()} or \lit{get_data()}. So:
\begin{python}
args = lib.get_args(['fixes'])
ds = lib.data.dataset.get_data(args)
\end{python}
`\lit{ds}' will only have data for the month of may. Convenient when testing a script on smaller data.

@@

With the information given above, we can automatically create a number of useful functions, detailled below.
All functions take as arguments \py{(args=None, **kwargs)} (see \titleref{sec:org-arg-processing}).
\paragraph{\texttt{get\_root}}
This simply returns the directory containing all data.

@@

\paragraph{\texttt{get\_data}}
This returns a dataset created with \hyperref{https://xarray.pydata.org/en/stable/generated/xarray.open_mfdataset.html}{\lit{xarray.open_mfdataset()}}
The files are supplied by the Finder object.
Other arguments to the function are supplied by the \lit{lib.data.}\-\lit{OPEN_MF_KW_DEF} dictionnary, completed by the argument \lit{open_mf_kw} of the \lit{lib.data.create_dataset} and \lit{lib.data.<dataset>.get_data} functions (see below).

@@

This function also takes a \lit{open_mf_kw} dictionnary argument, its members are passed to \lit{xr.open_mfdataset()} as keyword arguments.
Regarding the \lit{preprocess} argument, it can be useful to have a preprocessing function that take advantage of the information found in the filename (that can be retrieved by the Finder).
To setup such a function, special arguments in the dictionnary can be used:
\lit{preprocess_finder} should be a callable that takes in the XArray dataset to process, the corresponding filename, and the finder object, and it should return a XArray dataset.
The \lit{preprocess_finder_args} and \lit{preprocess_finder_kwargs} members are passed as additionnal arguments to this callable.

@@

\subsection{Metadata on creation}
It is useful to know when and how the datafiles originated.
To this end, the function \lit{lib.setup_metadata} can be used.
It adds to a dataset various global attributes:
\begin{itemize}
  \item \lit{"created_with"}: The name of the script, and the machine hostname
  \item \lit{"created_with_args"}: The args dictionnary as a string
  \item \lit{"created_on"}: The date and time of file creation
  \item \lit{"created_on_commit"}: The hash of the current commit/HEAD
\end{itemize}

>>>>

\lit{args} is a dictionary containing arguments name and their values.

\subsection{Default arguments}
The list of arguments names passed above (region, days, fixes) are from a list of default arguments.
Thoses are defined globally in \linkfile{lib/conf.ini}{lib/conf.ini}.
This file must be created; it can be copied from \linkfile{lib/conf_example.ini}{lib/conf_example.ini}.
Each default argument should have a name, type, and value. See the example conf file for more details.
The default values can be overridden when creating a dataset: see \titleref{sec:org-datasets}.

@@

\subsubsection{fixes}
Fixes are supplied as \lit{-fix <matcher> <string>}. They can then be found as a dictionary with the matchers name as keys. %chktex 44
See \titleref{sec:org-fixes} below.

@@

Now, we can do this automatically from the command line by supplying the appropriate argument: \shell{python script.py -fix m 05} (from the command line we can only supply a string, so it must match the date format). This will end up in the args dictionary, where it can be passed to \lit{get_finder()} or \lit{get_data()}. So:
\begin{python}
args = lib.get_args(['fixes'])
ds = lib.data.dataset.get_data(args)
\end{python}
`\lit{ds}' will only have data for the month of may. Convenient when testing a script on smaller data.

@@

With the information given above, we can automatically create a number of useful functions, detailed below.
All functions take as arguments \py{(args=None, **kwargs)} (see \titleref{sec:org-arg-processing}).
\paragraph{\texttt{get\_root}}
This simply returns the directory containing all data.

@@

\paragraph{\texttt{get\_data}}
This returns a dataset created with \hyperref{https://xarray.pydata.org/en/stable/generated/xarray.open_mfdataset.html}{\lit{xarray.open_mfdataset()}}
The files are supplied by the Finder object.
Other arguments to the function are supplied by the \lit{lib.data.}\-\lit{OPEN_MF_KW_DEF} dictionary, completed by the argument \lit{open_mf_kw} of the \lit{lib.data.create_dataset} and \lit{lib.data.<dataset>.get_data} functions (see below).

@@

This function also takes a \lit{open_mf_kw} dictionary argument, its members are passed to \lit{xr.open_mfdataset()} as keyword arguments.
Regarding the \lit{preprocess} argument, it can be useful to have a preprocessing function that take advantage of the information found in the filename (that can be retrieved by the Finder).
To setup such a function, special arguments in the dictionary can be used:
\lit{preprocess_finder} should be a callable that takes in the XArray dataset to process, the corresponding filename, and the finder object, and it should return a XArray dataset.
The \lit{preprocess_finder_args} and \lit{preprocess_finder_kwargs} members are passed as additional arguments to this callable.

@@

\subsection{Metadata on creation}
It is useful to know when and how the datafiles originated.
To this end, the function \lit{lib.setup_metadata} can be used.
It adds to a dataset various global attributes:
\begin{itemize}
  \item \lit{"created_with"}: The name of the script, and the machine hostname
  \item \lit{"created_with_args"}: The args dictionary as a string
  \item \lit{"created_on"}: The date and time of file creation
  \item \lit{"created_on_commit"}: The hash of the current commit/HEAD
\end{itemize}

### 2023-05-04 tex/article.tex
Fix spelling

Finally, the subpolar biome is characterized by a spring bloom, of which we measured the onset date both over fronts and in the background, for each year.
We defined the lag in bloom onset~\as{lag} as the bloom onset day over fronts minus the bloom onset day in the background.
Because the spring bloom onsets propagate from south to north, bloom onset days were inferred over latitudinal bands of limited width (\ang{5}).
We pooled apart front and non front pixels and computed the onset dates and their uncertainty based on the time series of the fronts and non-front \as{chl} median value.
We first filtered the \as{chl} median time series with a low-pass Butterworth filter of order~2 and cutting frequency \(1/20\,\text{days}^{-1}\).
The filtered time-series displayed strong variations in their phenology from year to year, but a bloom was always discernible.
We considered data from February to July, which allowed us to isolate the spring bloom and exclude the autumn bloom.
We detected the maximum value of \as{chl} in this time window, and defined the bloom onset as the time of maximum \as{chl} derivative prior to the time of maximum \as{chl}.
We defined the uncertainty in bloom onset date as the standard deviation of all days for which the \as{chl} derivative is above \pct{90} of its maximum value.
We defined the uncertainty in lag~\as{lag} as the square root of the sum of squared uncertainties in fronts and background.
Finally, we estimated the mean lag (over the 20~years of data and for each latitudinal band) as the weighted averaged lag, with the weights equal to the inverse of the uncertainties; and we estimated the mean lag uncertainty as the weighted standard deviation of all lag values around the mean lag, with the weights equal to the inverse of the uncertainties.
Our Eulerien approach relies on the hypothesis that the bloom evolves coherently in the background (resp.\ over fronts) within each latitudinal band, which is suggested by high-resolution models of the bloom (e.g.\ \cite{levy_2005a,karleskind_2011}).
It is imperfect as the bloom evolves along Lagrangian trajectories, but provided very consistent results.

@@

The fraction of the area occupied by fronts also varies with seasons, with generally fewer fronts in summer (\nref{fig:ts-climato}d-f).
In the permanent subtropical biome, weak fronts cover on average up to \pct{12} in spring and drop to \pct{2} in summer.
In the seasonal subtropical biome, the variation is from \pct{27} to \pct{13} for weak fronts, and \pct{8} to \pct{4} for strong fronts.
In the subpolar biome, strong fronts cover between \pct{11} in summer and \pct{26} at their peak.
This seasonality is consistent with other estimates that submesoscale frontal activity is greater in winter due to greater mixed-layer thickness compared to summer (\cite{callies_2015}), and also consistent with the most recent modelling results by \textcite{dong_2020} in another Western Boundary system (the Kuroshio Extension) where they show that the strongest submesoscale dynamics occur with a lag of about a month after the mixed layer thickness maximum is reached.

@@

Our evaluation of the regional surplus in \as{chl} associated with fronts assumes that the local \as{chl} excess due to fronts is only located over fronts.
However, the localization of the highest \as{chl} do not perfectly coincide with elevated values of HI;\ they are places where the HI is large and \as{chl} is small, and there are also places with elevated patches of \as{chl} outside HI contours (see for example in \nref{fig:zoom}).
The apparent mismatch between the exact location of fronts at a given time and the exact location of the \as{chl} response to frontal dynamics at the same time may be due to different factors, such as the very dynamic nature of fronts which can lead to the chaotic advection of phytoplankton (and/or nutrients) outside of fronts (\cite{abraham_1998}), the time-scale needed for phytoplankton to respond to nutrient supplies at fronts, or trophic interactions at the front (as shown, for instance, in the modelling study of \textcite{mangolte_2022}).
This implies that part of the excess \as{chl} attributable to frontal activity may actually be located outside of fronts. Our results show that the median distributions of \as{chl} significantly differ over fronts and outside of fronts, which suggests that most of the signal due to fronts actually occurs over fronts, nevertheless our estimate of the regional impact of fronts is likely an underestimate due to the signal located outside of fronts.

@@

Finally, an important outcome of this study is that the surplus in phytoplankton biomass associated with fronts can be stronger in blooming biomes than in oligotrophic ones.
In fact in regions where nutrients are not limiting productivity, fronts have been shown to subduct excess nutrients (\cite{oschlies_2002, gruber_2011}) and excess biomass (\cite{lathuiliere_2010}), rather than to lead to an increase in biomass as in oligotrophic regions.
This effect of a decreased biomass, suggestive of subduction, is observed here in the northern parts of the subpolar biome between May and September (\nref{fig:latbands-n}b-d).
But we also find that in the subpolar biome, the \as{chl} excess over fronts can reach \pct{150} during the bloom and \pct{50} during summer (\nref{fig:latbands-n}), while in the permanent subtropical biome it never exceeds \pct{10} (\nref{fig:latbands-s}).
The enhancement of the spring bloom by submesoscale frontal processes observed here was recently also put forward in a modelling study by \textcite{simoes-sousa_2022}.

@@

The method that we used to quantify differences in bloom timing over fronts and background is based on the time evolution of an eulerian quantity, the \as{chl} median over latitudinal bands, whereas the bloom evolves along lagrangian trajectories.
Considering a rather small area, as we have done here, is a way of overcoming the difficulty of following the temporal evolution on fronts whose life history is too complex to be captured and shorter than the bloom itself.
It also limits the impact that the northward propagation of the bloom could have on the temporal assessment.
It should also be noted that it is inherently difficult to pinpoint the precise onset and end days of a bloom, as the spring bloom shows large intraseasonal variability in its characteristics; its beginning can be more or less sudden, and is often made of multiple peaks (\cite{keerthi_2020}).

@@

Although limited to the Gulf Stream region, this study provides a well-tested methodology that could enable the study of the links between small-scale ocean physics and phytoplankton response in other regions of the global ocean.
In addition, these results on the importance of fronts for phytoplankton biomass and phenology could also be used to evaluate models coupling ocean physics and phytoplankton at high spatial resolution, or to test parameterizations representing the effect of small scales on phytoplankton production in coarser resolution models.
Finally, the combination of these observation-based results with theoretical arguments and well-assessed models should also allow us to better constrain the response of phytoplankton production to climate change (\cite{couespel_2021}), which still has very large uncertainties as shown by the latest set of Earth system models (\cite{kwiatkowski_2020}).

@@

\begin{articleSubBlock}{Acknowledgements}
  CH benefited from a PhD scholarship by ENS.\@ The project was supported by TOSCA CNES and by the ENS CHANEL chair.
  We thank Daniele Iudicone, Francesco d'Ovidio, Sakina-Dorothée Ayata, and Amala Mahadevan for the useful discussions which helped to refine our methodology. We thank Xioa Liu for helping us reproduce their work.

>>>>

Finally, the subpolar biome is characterized by a spring bloom, of which we measured the onset date both over fronts and in the background, for each year.
We defined the lag in bloom onset~\as{lag} as the bloom onset day over fronts minus the bloom onset day in the background.
Because the spring bloom onsets propagate from south to north, bloom onset days were inferred over latitudinal bands of limited width (\ang{5}).
We pooled apart front and non front pixels and computed the onset dates and their uncertainty based on the time series of the fronts and non-front \as{chl} median value.
We first filtered the \as{chl} median time series with a low-pass Butterworth filter of order~2 and cutting frequency \(1/20\,\text{days}^{-1}\).
The filtered time-series displayed strong variations in their phenology from year to year, but a bloom was always discernible.
We considered data from February to July, which allowed us to isolate the spring bloom and exclude the autumn bloom.
We detected the maximum value of \as{chl} in this time window, and defined the bloom onset as the time of maximum \as{chl} derivative prior to the time of maximum \as{chl}.
We defined the uncertainty in bloom onset date as the standard deviation of all days for which the \as{chl} derivative is above \pct{90} of its maximum value.
We defined the uncertainty in lag~\as{lag} as the square root of the sum of squared uncertainties in fronts and background.
Finally, we estimated the mean lag (over the 20~years of data and for each latitudinal band) as the weighted averaged lag, with the weights equal to the inverse of the uncertainties; and we estimated the mean lag uncertainty as the weighted standard deviation of all lag values around the mean lag, with the weights equal to the inverse of the uncertainties.
Our Eulerian approach relies on the hypothesis that the bloom evolves coherently in the background (resp.\ over fronts) within each latitudinal band, which is suggested by high-resolution models of the bloom (e.g.\ \cite{levy_2005a,karleskind_2011}).
It is imperfect as the bloom evolves along Lagrangian trajectories, but provided very consistent results.

@@

The fraction of the area occupied by fronts also varies with seasons, with generally fewer fronts in summer (\nref{fig:ts-climato}d-f).
In the permanent subtropical biome, weak fronts cover on average up to \pct{12} in spring and drop to \pct{2} in summer.
In the seasonal subtropical biome, the variation is from \pct{27} to \pct{13} for weak fronts, and \pct{8} to \pct{4} for strong fronts.
In the subpolar biome, strong fronts cover between \pct{11} in summer and \pct{26} at their peak.
This seasonality is consistent with other estimates that submesoscale frontal activity is greater in winter due to greater mixed-layer thickness compared to summer (\cite{callies_2015}), and also consistent with the most recent modeling results by \textcite{dong_2020} in another Western Boundary system (the Kuroshio Extension) where they show that the strongest submesoscale dynamics occur with a lag of about a month after the mixed layer thickness maximum is reached.

@@

Our evaluation of the regional surplus in \as{chl} associated with fronts assumes that the local \as{chl} excess due to fronts is only located over fronts.
However, the localization of the highest \as{chl} do not perfectly coincide with elevated values of HI;\ they are places where the HI is large and \as{chl} is small, and there are also places with elevated patches of \as{chl} outside HI contours (see for example in \nref{fig:zoom}).
The apparent mismatch between the exact location of fronts at a given time and the exact location of the \as{chl} response to frontal dynamics at the same time may be due to different factors, such as the very dynamic nature of fronts which can lead to the chaotic advection of phytoplankton (and/or nutrients) outside of fronts (\cite{abraham_1998}), the time-scale needed for phytoplankton to respond to nutrient supplies at fronts, or trophic interactions at the front (as shown, for instance, in the modeling study of \textcite{mangolte_2022}).
This implies that part of the excess \as{chl} attributable to frontal activity may actually be located outside of fronts. Our results show that the median distributions of \as{chl} significantly differ over fronts and outside of fronts, which suggests that most of the signal due to fronts actually occurs over fronts, nevertheless our estimate of the regional impact of fronts is likely an underestimate due to the signal located outside of fronts.

@@

Finally, an important outcome of this study is that the surplus in phytoplankton biomass associated with fronts can be stronger in blooming biomes than in oligotrophic ones.
In fact in regions where nutrients are not limiting productivity, fronts have been shown to subduct excess nutrients (\cite{oschlies_2002, gruber_2011}) and excess biomass (\cite{lathuiliere_2010}), rather than to lead to an increase in biomass as in oligotrophic regions.
This effect of a decreased biomass, suggestive of subduction, is observed here in the northern parts of the subpolar biome between May and September (\nref{fig:latbands-n}b-d).
But we also find that in the subpolar biome, the \as{chl} excess over fronts can reach \pct{150} during the bloom and \pct{50} during summer (\nref{fig:latbands-n}), while in the permanent subtropical biome it never exceeds \pct{10} (\nref{fig:latbands-s}).
The enhancement of the spring bloom by submesoscale frontal processes observed here was recently also put forward in a modeling study by \textcite{simoes-sousa_2022}.

@@

The method that we used to quantify differences in bloom timing over fronts and background is based on the time evolution of an Eulerian quantity, the \as{chl} median over latitudinal bands, whereas the bloom evolves along Lagrangian trajectories.
Considering a rather small area, as we have done here, is a way of overcoming the difficulty of following the temporal evolution on fronts whose life history is too complex to be captured and shorter than the bloom itself.
It also limits the impact that the northward propagation of the bloom could have on the temporal assessment.
It should also be noted that it is inherently difficult to pinpoint the precise onset and end days of a bloom, as the spring bloom shows large intraseasonal variability in its characteristics; its beginning can be more or less sudden, and is often made of multiple peaks (\cite{keerthi_2020}).

@@

Although limited to the Gulf Stream region, this study provides a well-tested methodology that could enable the study of the links between small-scale ocean physics and phytoplankton response in other regions of the global ocean.
In addition, these results on the importance of fronts for phytoplankton biomass and phenology could also be used to evaluate models coupling ocean physics and phytoplankton at high spatial resolution, or to test parametrization representing the effect of small scales on phytoplankton production in coarser resolution models.
Finally, the combination of these observation-based results with theoretical arguments and well-assessed models should also allow us to better constrain the response of phytoplankton production to climate change (\cite{couespel_2021}), which still has very large uncertainties as shown by the latest set of Earth system models (\cite{kwiatkowski_2020}).

@@

\begin{articleSubBlock}{Acknowledgments}
  CH benefited from a PhD scholarship by ENS.\@ The project was supported by TOSCA CNES and by the ENS CHANEL chair.
  We thank Daniele Iudicone, Francesco d'Ovidio, Sakina-Dorothée Ayata, and Amala Mahadevan for the useful discussions which helped to refine our methodology. We thank Xioa Liu for helping us reproduce their work.

### 2023-05-04 tex/biblio.tex
Fix spelling

J'ai été amené à utiliser durant ma thèse de nombreux outils informatiques sans lesquels ce travail n'aurait pu aboutir, ou en tout cas avec assurément beaucoup plus de difficultés.
Il me semble important d'en citer au moins une partie ici.
La reproductibilité de mon travail est déjà garantie par la mise à disposition de mes codes (accompagnés d'une courte documentation et de tout le nécessaire pour reproduire les résultats) sur un dépôt public\footnote{%
  Dépot Gitlab: \glsurl{gitlab}}
et dont une version est également déposée sur un répertoire Zenodo\footnote{\glsurl{zenodo}}.
Il s'agit ici plutôt de créditer les nombreux·ses contributeurs et contributrices qui ont participé à l'élaboration de ces outils.

Bien évidemment beaucoup des calculs reposent sur des librairies qui ne sont plus à présenter: \citesoft{numpy}, \citesoft{scipy}, et \citesoft{pandas}.
Cependant il est plus commun (en géosciences en tout cas) d'interfacer avec la librairie \citesoft{xarray}.
Les outils de \citesoft{dask} se sont également avérés indispensables pour gérer les quantités importantes de données que sont les notres.

Il convient de citer plusieurs paquets permettant le calcul efficace d'histogrammes:
\citesoft{xhistogram} qui implémente ses fonctionalités à partir de fonctions NumPy et Dask \enquote*{élémentaires}, et
\citesoft{dask-histogram} qui s'appuie sur la librairie C \citesoft{boost}.

@@

Enfin il me parait approprié de citer les outils \citesoft{ipython} avec lequel j'execute tous mes scripts, \citesoft{mamba} et le projet \citesoft{conda-forge} lesquels me permettent de gérer les environments Python, et enfin le travail épatant derrière le framework de configuration \citesoft{doom} sur lequel tous les scripts et ce manuscrit (entre autres) ont été écrits.

>>>>

J'ai été amené à utiliser durant ma thèse de nombreux outils informatiques sans lesquels ce travail n'aurait pu aboutir, ou en tout cas avec assurément beaucoup plus de difficultés.
Il me semble important d'en citer au moins une partie ici.
La reproductibilité de mon travail est déjà garantie par la mise à disposition de mes codes (accompagnés d'une courte documentation et de tout le nécessaire pour reproduire les résultats) sur un dépôt public\footnote{%
  Dépôt Gitlab: \glsurl{gitlab}}
et dont une version est également déposée sur un répertoire Zenodo\footnote{\glsurl{zenodo}}.
Il s'agit ici plutôt de créditer les nombreux·ses contributeurs et contributrices qui ont participé à l'élaboration de ces outils.

Bien évidemment beaucoup des calculs reposent sur des librairies qui ne sont plus à présenter: \citesoft{numpy}, \citesoft{scipy}, et \citesoft{pandas}.
Cependant il est plus commun (en géosciences en tout cas) d'interfacer avec la librairie \citesoft{xarray}.
Les outils de \citesoft{dask} se sont également avérés indispensables pour gérer les quantités importantes de données que sont les nôtres.

Il convient de citer plusieurs paquets permettant le calcul efficace d'histogrammes:
\citesoft{xhistogram} qui implémente ses fonctionnalités à partir de fonctions NumPy et Dask \enquote*{élémentaires}, et
\citesoft{dask-histogram} qui s'appuie sur la librairie C \citesoft{boost}.

@@

Enfin il me parait approprié de citer les outils \citesoft{ipython} avec lequel j’exécute tous mes scripts, \citesoft{mamba} et le projet \citesoft{conda-forge} lesquels me permettent de gérer les environnements Python, et enfin le travail épatant derrière le framework de configuration \citesoft{doom} sur lequel tous les scripts et ce manuscrit (entre autres) ont été écrits.

### 2023-05-04 tex/conclusion.tex
Fix spelling

Au delà de ces résultats, nous incrémentons et validons une méthode de détection des fronts dans une zone complexe.
En particulier, nous montrons son adéquation à quantifier l'intensité des fronts détectés.
En outre, nous reconnaissons que malgré l'importance et les applications potentielles de la détection de fronts, il n'existe pas encore d'outils ou de produits facilitant leur usage scientifique.
Sans pour autant fournir une solution définitive à cette lacune, ce travail souligne des objectifs à atteindre et fournit un exemple d'outil facilement réutilisable.
Nous avons également pu identifier un certains nombre de difficultés se dressant devant l'établissement de tels outils ou produits.
De multiples algorithmes de détection de fronts existent déjà, cependant leur usage n'est pas forcément trivial.
En cause est leur distribution dans des canaux et des languages inadaptés, ou manquant de documentation, et dont on ne peut que souhaiter qu'elle se rapproche des standards et outils aujourd'hui en place dans le milieu de l'open-source.
Ces manquements incombent un travail supplémentaire à leur application, pourtant déjà complexe du fait, entre autres, de la nécessité de s'adapter aux caractéristiques régionales (que nous montrons ici), et du large volume de données nécessaire aux évaluations de l'impact des fines échelles.

>>>>

Au delà de ces résultats, nous incrémentons et validons une méthode de détection des fronts dans une zone complexe.
En particulier, nous montrons son adéquation à quantifier l'intensité des fronts détectés.
En outre, nous reconnaissons que malgré l'importance et les applications potentielles de la détection de fronts, il n'existe pas encore d'outils ou de produits facilitant leur usage scientifique.
Sans pour autant fournir une solution définitive à cette lacune, ce travail souligne des objectifs à atteindre et fournit un exemple d'outil facilement réutilisable.
Nous avons également pu identifier un certains nombre de difficultés se dressant devant l'établissement de tels outils ou produits.
De multiples algorithmes de détection de fronts existent déjà, cependant leur usage n'est pas forcément trivial.
En cause est leur distribution dans des canaux et des langages inadaptés, ou manquant de documentation, et dont on ne peut que souhaiter qu'elle se rapproche des standards et outils aujourd'hui en place dans le milieu de l'open-source.
Ces manquements incombent un travail supplémentaire à leur application, pourtant déjà complexe du fait, entre autres, de la nécessité de s'adapter aux caractéristiques régionales (que nous montrons ici), et du large volume de données nécessaire aux évaluations de l'impact des fines échelles.
